#Week 6 analysis of longitudinal data

#introduce datasets
#RATS includes weight data of rats with 3 different treatments.
#BPRS includes weekly bprs phychological test scores for me with 2 treatments.

#quick summary of data
```{r}

#read the data
BPRS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BPRS.txt", header = T)
str(BPRS)
#treatment, subject and 0 to 8 weeks of BPRS summed scores. 
summary(BPRS)

RATS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", header = T)
str(RATS)
#ID and group, and then weight data from 11 time intervals from day 1 to day 64.
summary(RATS)

#factorization
BPRS$treatment <- as.factor(BPRS$treatment)
BPRS$subject <- as.factor(BPRS$subject)
RATS$ID <- as.factor(RATS$ID)
RATS$Group <- as.factor(RATS$Group)

#convert to long form
library(dplyr)
library(tidyr)
BPRSL <-  BPRS %>% gather(key = weeks, value = bprs, -treatment, -subject)
BPRSL <-  BPRSL %>% mutate(week = as.integer(substr(weeks, 5,5)))
BPRSL <- subset(BPRSL, select = -weeks)
str(BPRSL)
colnames(BPRSL)
summary(BPRSL)


RATSL <- RATS %>%
  gather(key = WD, value = Weight, -ID, -Group) %>%
  mutate(Time = as.integer(substr(WD,3,4)))
str(RATSL)
colnames(RATSL)
summary(RATSL)
```
graphical summary
```{r}
library(ggplot2)
ggplot(RATSL, aes(x = Time, y = Weight, group = ID)) +
  geom_line(aes(linetype = Group)) +
  scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 10)) +
  scale_y_continuous(name = "Weight (grams)")+
  theme(legend.position = "top")
```
Starting weigths of treatment 1 seem quite a bit lower compared to treatments 2 and 3.
To understand the change in variation across time, we can plot standardised values:
```{r}
library(tidyr)
library(dplyr)
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate(stdWeight = (Weight - mean(Weight))/sd(Weight)) %>%
  ungroup()
ggplot(RATSL, aes(x = Time, y = stdWeight, group = ID)) +
  geom_line(aes(linetype = Group)) +
  scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 10)) +
  scale_y_continuous(name = "Standardised weight")+
  theme(legend.position = "top")
```
Variability seems rather constant with time.
Next summary graph of the mean weight for each group
```{r}
n <- RATSL$Time %>% unique() %>% length()
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(Weight), se = sd(Weight)/sqrt(n) ) %>%
  ungroup()
ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
 
  geom_point(size=3) +

  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.8,0.8)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
```
Boxplot summary
```{r}
RATSL1 <- RATSL %>%
  filter(Time > 1) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()
ggplot(RATSL1, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight)")

```
Filter 3 outliers out
```{r}

RATSL2 <- filter(RATSL1, Group == "1" & mean >250 |Group == "2" | Group == "3")
RATSL3 <- filter(RATSL2, Group == "2" & mean < 550  |Group == "1" | Group == "3")
RATSL4 <- filter(RATSL3, Group == "3" & mean > 500  |Group == "1" | Group == "2")
ggplot(RATSL4, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight)")
```
Anova with no baseline
```{r}

```

